---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: "Ashton Cloer"
date: "Fall 2023"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
#1
library(tidyverse)
library(agricolae)
library(here)
library(ggthemes)
library(tidyr)
getwd()
here()

NTL_LTER_Lake_ChemistryPhysics <- 
  read.csv(
    here(
    "EDE_Fall2023/Data/Raw/NTL_LTER_Lake_ChemistryPhysics_Raw.csv"),
    stringsAsFactors = TRUE)

NTL_LTER_Lake_ChemistryPhysics$sampledate <- 
  mdy(
  NTL_LTER_Lake_ChemistryPhysics$sampledate)

class(
  NTL_LTER_Lake_ChemistryPhysics$sampledate)

#2
mytheme1 <- theme_gray(
  base_size = 14, 
  base_family = "Times") +
  theme(
    axis.text = element_text(
      color = "gray"), 
        legend.position = "bottom")
theme_set(mytheme1)
```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature in July does not change with depth across all lakes
Ha: Mean lake temperature in July does change with depth across all lakes


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4

NTL_LTER_wrangled <- 
  NTL_LTER_Lake_ChemistryPhysics %>%
  mutate_at(vars(sampledate), 
            funs(year, 
                 month, day)) %>%
  filter(month == '7') %>%
  select('lakename', 'year4', 
         'daynum', 'depth', 
         'temperature_C') %>%
  na.omit()
  
#5
temp_by_depth <- 
  ggplot(NTL_LTER_wrangled, 
         aes(x = depth, y = temperature_C)) +
  geom_point() +
  geom_smooth(
    method = 'lm', col="blue") + 
  ylim(0,35) +
  labs(x = "Depth", 
       y = "Temperature (C)") +
  ggtitle(
    "How Temperature Varies by Depth in July") +
  mytheme1
print(temp_by_depth)
```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The figure suggests that temperature decreases as depth increases. The distribution of points in the graph suggest the trend follows either a negative linear, or slightly negative logarithmic pattern, given how the steepness of the slope increases towards the middle of the distribution and tapers off towards the highest and lowest depths.   


7. Perform a linear regression to test the relationship and display the results

```{r linear.regression}
#7
lm_temp_by_depth <-
  lm(
    data = NTL_LTER_wrangled, 
    temperature_C ~ depth)
summary(lm_temp_by_depth) 
```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The model suggests that depth and temperature have a negative relationship, where temperature decreases as depth increases. Changes in depth was found to explain 73.87% of the variability in temperature, degrees of freedom were calculated to 9726 and the p-value was 2.2e-16 or less than 0. For each 1m change in depth, temperature was found to change by 1.94 degrees.  

---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model}
#9
tempAIC <- lm(
  data = NTL_LTER_wrangled, 
  temperature_C ~ depth + year4 + daynum)
step(tempAIC)

#10
temp_model <- lm(
  data = NTL_LTER_wrangled, temperature_C ~ depth + year4 + daynum) 
summary(temp_model)
```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: The final set of explanatory variables to predict temperatures is depth, year4, and daynum. This model explains 74.12% of the observed variance in temperature. This is an improvement over the model using just depth which only explains 73.87% of the observed variance in temperature. 



---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model}
#12
anova_temp_by_Depth <- aov(
  data = NTL_LTER_wrangled, 
  temperature_C ~ lakename)
summary(anova_temp_by_Depth)

anova2_temp_by_Depth <- lm(
  data = NTL_LTER_wrangled, 
  temperature_C ~ lakename)
summary(anova2_temp_by_Depth)
```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: Based on the anova test (p value < 0 or 2e-16), we can reject the null hypothesis and suggest that there is a difference in mean temperatures amongst the lake locations in the month of July. 



14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
temp_by_depth_bylake <- 
  ggplot(
    NTL_LTER_wrangled, 
    aes(x = depth, 
        y = temperature_C, 
        color = lakename)) +
  geom_point(alpha = 0.5) + 
  geom_smooth(
    method = 'lm', 
    col="blue") + 
  ylim(0,35) +
  labs(x = "Depth", 
       y = "Temperature (C)") +
  ggtitle(
    "How Temperature Varies by Depth in July at Various Locations") +
  mytheme1
print(temp_by_depth_bylake)

```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test}
#15
Tukey.HSD<- TukeyHSD(anova_temp_by_Depth)
HSD_test<- HSD.test(
  anova_temp_by_Depth, 
  "lakename", 
  group = TRUE)
print(HSD_test)
print(Tukey.HSD)
```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer: Ward Lake (p value = 0.782) seems to have the same mean temperature as Peter Lake, statistically speaking, given the difference in means of 1.142 is supported by a p-value that is unsignficant. Central Long lake, Tuesday Lake, and East Long lake seem to have means that are statiscally distinct from all other lakes. 

 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: HSD.test and or a two sample t-test



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
NTL_LTER_wrangled2 <- 
  NTL_LTER_wrangled %>%
  filter(
    lakename %in% c(
      "Crampton Lake",
      "Ward Lake"))

lake.twosample <- t.test(
  NTL_LTER_wrangled2$temperature_C ~ 
    NTL_LTER_wrangled2$lakename)
lake.twosample

lake.twosample2 <- lm(
  NTL_LTER_wrangled2$temperature_C ~ 
    NTL_LTER_wrangled2$lakename)
summary(lake.twosample2)
```

>Answer: According to the two sample t-test, the mean temperatures for Crampton and Ward Lake are not statiscally different (p = 0.2649). Therefore, based on the ananlysis, we cannot reject the null hypothesis which states that the true difference in means between group Crampton Lake and Ward Lake is equal to zero. This matches my answer to question 16. 
